---
title: "reading data into R answers"
author: "Ziyi Kang"
date: "Friday, January 23, 2015"
output: html_document
---

Note: Statements that are Commonly used for reading data into R
```{r,eval=FALSE} 
-scan()
-read.table()
-read.fwf()
-read.csv()
-read.csv2()
-read.delim()
read.delim2()
```
Please download the dataset sample and codebook from National Survey of Family Growth and take a quick look on what is recorded in the code book.

-Female Pregnancy Data File (2002FemPreg.dat) http://www.cdc.gov/nchs/nsfg/nsfg_cycle6.htm

-Pregnancy Codebook(PDF): http://www.cdc.gov/nchs/nsfg/nsfg_cycle6.htm


Task 1:
Read the first 10 column names into R
```{r}
FemPreg.colnames<-scan(what="")
```

Task 2:
Read the first 10 categories of data into R from .dat file

```{r}
FemPreg.colnames<-c("CASEID","pregordr","HOWPREG_N","HOWPREG_P","moscurrp","NOWPRGDK","PREGEND1","PREGEND2","NBRNALIV","MULTBRTH")

FemPreg.colwidth<-c(12,2,2,1,1,1,1,1,1,1)

FemPreg<-read.fwf("C:/Users/Ziyi/Desktop/2002FemPreg.dat",widths=FemPreg.colwidth,col.names=FemPreg.colnames)

head(FemPreg)
```

Task 3:
Export different format of data(.dat .csv .txt .xls) and try to figure out the meaning of all the arguments in the statement. 
```{r}
#Creat .dat .csv .txt files of the dataset
write.csv(FemPreg,"FemPreg2002.csv",row.names=FALSE)
write.table(FemPreg,"FemPreg2002.txt",row.names=FALSE)
```

Task 4
read .txt files into R and try to change from the default setting of the argument
```{r}
FemPreg1<-read.table("FemPreg2002.txt",header=TRUE,sep="")
head(FemPreg1)
```
