---
title: 'Lecture 11: Transformations and Modeling'
author: "IE 0015"
date: "March 23, 2017"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Transforming data
============================

- We transform data because it allows us to more clearly identify patterns in the data.
- We think there is a relationship, a transform allows us to see if the relationship is strong.
- Many numerical methods (e.g. linear regression) can work better if the model is transformed.

Some transforms you already know
===========================

- Filtering - Examine a specific part of the dataset or remove outlier cases.
  - Or filter to look at outliers more closely.


```{r, echo=FALSE}
library(ggplot2, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
```
```{r}
data(diamonds)
nrow(diamonds)
ggplot(diamonds, aes(x,y)) + geom_bin2d() 
```

```{r}
diamonds_ok <- filter(diamonds, x>0, y>0, y < 20)
ggplot(diamonds_ok) + geom_bin2d(aes(x,y)) + geom_abline(slope=1, color="white", alpha=0.5)
```

Logical operators for filters
===========

- `x==y`: x and y are equal.
- `x!=y`: x and y are not equal.
- `x %in% c("a", "b", "c")`: x is one of the values in the set right hand side.
- `x > y, x >= y, x < y, x <= y`: greater than, greater than or equal to, less than, less than or equal to.



And combine them with logical operators:
======

- `!x` - Not x, flips TRUE and FALSE
- `x&y`: TRUE if both x and y are TRUE.
- `x|y`: TRUE if either x or y (or both) are TRUE.
- `xor(x,y)`: TRUE if either x or y are TRUE, but not both (exclusive or).

Transforming data
=================

- Some typical transformations

- Log transforms - turns multiplicative (compounding) relationships into additive relationships.
  - Often used to compress data that ranges over several orders of magnitude.
- Relative difference - Use log(x/y) instead of the actual values x and y. (note: this is better than x/y because it is symmetric, the order of x and y does not matter)
- Integrating or differentiating to transform variables.  E.g. transform distance and time to speed or acceleration.
- Cartesian (rectangular) to/from polar coordinates
- Centering and scaling 
  - Converting to standard normal subtracting from a mean and dividing by standard deviation
  - Converting point to a percentile of the entire range
  
Rescale example from last lecture
==========

```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(0, 5, 10))
```

Note: statistical software and rescaling
==============

- R, SAS, SPSS do this automatically in their statistical routines.
- For other statistical software, the accuracy of results may be improved if you do this.


In dplyr
============

- Use group_by and summarize to calculate needed summary statistics
- Use mutate to do the transformation

Use diamonds example
==============

Show how summary statistics can be used to explain odd data.

Modeling
===========

The Art of Modeling
===================

> -  We use models in an attempt to gain insights about some aspects of the real world.
> -  Assumes *a priori* the existence of *ground truth* which an impartial and omniscient observers would agree upon.
> -  We start by deciding what aspect of the real world we are interested in.
  -  Narrow the focus of the model.

Some definitions
============

> -  *System* is a set of elements that interact or interrelate in some fashion.
> -  *Entities* are elements that make up the system
> -  If an element does not interact or interrelate with other elements of the system, it is irrelevant to our goal.
> -  A system may be made up of *sub-systems*, and may be a sub-system for another system.

What is a model
==========

-  A *model* is a system we use to study in place of another system.
-  The cost of building and studying a model is often a small fraction of the cost of experimenting with the real system.
-  Benefits
	-  Mitigate risk
	-  Cause events to occur on demand to develop experience

Types of models
==========

-  Physical models - Wind tunnels.  Concept cars.
-  Mechanical models - Anatomy models for medicine.
-  Mathematical equations - e.g. trajectory of a falling object.
-  *Computer simulation* - A model that is implemented as a computer program.

Purpose of models
===========

> -  To mimic or describe the behavior of the system being modeled.
> -  It can be *simplified* or *abstracted*
> -  It should describe the *essential* characteristics of the system.
> -  We evaluate the model on how well its *outputs* match the observed outputs of reality.
> -  Then, we want to see how the model responds when we make changes.
> -  So we want to *observe* outputs and be able to *control* the model through inputs.
> -  The model is thus created to address a specific question.
> -  Insight, not numbers

Models for inference
=========

\begin{figure}
\includegraphics[width = 0.6\textwidth]{figures/modelinsight}
\caption{A model yields insights and inferences}
\end{figure}


Modeling for data analysis
===============

- Remove the so-what factor
- Identify effects beyond what is known

Diamond example
===========

-  Look at plot of carat vs price
-  Is this linear
> - Look at the distributions on both axis and the relationship?
> - What would make it linear?


```{r}
dplot <- ggplot(diamonds)
dplot + geom_point(aes(x=carat, y=price))
```

Log-log plot
===============

- Plot with log of carat for x axis, log of price for y axis

```{r}
dplot + geom_point(aes(x=log(carat), y=log(price)))
```

Using dplyr (if the transform was not this easy)
=================

- Add in a linear model (lm) to the curve
```{r}
diamonds2<-diamonds %>% 
  mutate(lcarat = log2(carat), lprice = log2(price))
ggplot(diamonds2, aes(x=lcarat, y=lprice)) + 
  geom_bin2d() +
  geom_smooth(method="lm", se=FALSE, size=2, color="yellow")
```

Linear modeling
============

- The other use of models is to account for aspects that you already know are true, but you want to examine the influence of other variables.
-  Let's look at clarity
```
a measurement of how clear the diamond is (I1 (worst), SI1, SI2, VS1, VS2, VVS1, VVS2, IF (best))
```
```{r}
dplot+geom_boxplot(aes(x=clarity, y=price))
```

Removing the influence of known predictors
==============================

-  Condition on another variable (carat)
-  In design of experiments, this is called *blocking*
-  We use linear regression models to analyze this type of experiment.

Experimental design
===================

Randomized experiments are built on four principles. If we cannot make a perfect experiment, we have to think about how violating these principles effects our analysis of data.

- Controlling - We assign observations to treatment groups, and try to control for differences in groups.
- Randomization - Randomize assignments of observations to treatment groups.
- Replication - Observe more cases so that we can more accurately estimate the effect of explanatory variables on response.
- Blocking - If we suspect that variables, other than the treatment, influence the response, we may group observations by this variable and then randomize assignment of these observations to treatment groups.

Linear models
================

- Multiple linear regression
- Response (dependent) variable
- Explanatory (independent, treatment) variable
- Blocking (independent, but not treatment)

$$y = Ax + b$$

Implementing linear models in R
============

- `lm` function
- Takes a model as input
  - `response ~ explanatory variables`
  - Left unsaid is a constant term
  - `response ~ explanatory variables - constant`

```{r}
lmcarat <- lm(lprice ~ lcarat, diamonds2)
summary(lmcarat)
```

Visualizing models
===========

-  Look at clarity
```{r}
dplot+geom_boxplot(aes(x=clarity, y=price))
```

Account for carat
==================

```{r}
lmclarity <- lm(lprice ~ lcarat + clarity, diamonds2)
summary(lmclarity)
```
Model summaries
=============

-  Look at points that do not fit the model
-  glance

Ordering based on model fit using reorder
===============

Ordering based on fit
===============