---
title: "A - IE 0015 Spring 2015 Final - A"
author: "IE 0015 Information Systems Engineering"
output: pdf_document
---

```{r, echo=FALSE}
options(digits=1)
library(ggplot2)
library(faraway)
```

1.  There are four questions and an article for this exam.
2.  You may have a double sided, hand-written page of notes (crib sheet)
3.  You may have a stand-alone calculator (e.g. not your cell phone). If you do not have a calculator, when you have a calculation you may set up the calculation and provide an approximation of the answer for full credit.
4.  Answer questions in the blue book provided. Additional blue books may be available if needed.
5.  You have 2 hours for the exam.

# 1.  Mobile phones in sub-saharan Africa (20 pts)

Mobile Phones Are Revolutionizing Personal Finance In Sub-Saharan Africa

Discuss the article in terms of:

A.  A CoNVO Statement and 
B.  Data Volume, Velocity, Variety


ANSWER
--------

-  Context:  Identify organization. Identify mission.  Mission is a broader context than the project.
-  Need:  Identify a decision to be made or a problem to be resolved.  Identify a question that can be answered with data.
-  Vision:  A description of a presentation of data that would answer the question identified in the Need.
-  Outcome:  A description of how the organization identified in the Context could respond when given the presentation described in the Vision.
-  Volume: How much data is there
-  Velocity: How do subjects change over time
-  Variety: What data is available on every subject


#  2.  Physician complaints (35 points)

Table 1 contains data for 20 emergency room doctors at an Australian hospital regarding complaints.  We are interested in the rate of complaints for doctors, however, the records for three of the doctors is incomplete because we do not have the number of patient visits handled by that doctor.

- visits - the number of patient visits
- complaints - the number of complaints
-  residency - is the doctor in residency training N or Y
- gender - gender of doctor F or M
- revenue - dollars per hour earned by the doctor
- hours - total number of hours worked

```{r, echo=FALSE}
data(esdcomp)
esdfinal <- esdcomp[1:20, ]
esdfinal$visits[c(2, 10, 14)] = NA
esdfinal
#gender.color = c(M="gray0", F="gray1")
ggplot(esdfinal, aes(y=complaints, x=revenue, shape= residency, color=gender)) + geom_point() + geom_jitter() + scale_color_grey(start=0.0, end=0.5) + scale_size_continuous(range = c(3,4)) + theme_bw() + ggtitle(label = "Complaints by dollars per hour earned by the doctor")
```

A.  One of your colleagues thinks that the best thing to do is use all available data and run a linear regression using the other five data elements available to fill in the number of visits. Assume that a linear regression is the correct method to use to impute missing values in this case.  Argue either in favor or against this argument (Note: an argument against should discuss what should be removed from a linear regression and why.)

B.  Assume that the variables that are important are residency, gender, and hours worked.  Use kNN imputation to impute the number of visits for doctor 2. Use $k=2$

C.  Assume that you can categorize on residency and gender.  Use mean value imputation to impute the number of visits for doctor 2.

D.  Discuss how you would choose between the methods for imputing missing value:  mean value, random selection, kNN, and linear regression.

ANSWER
------

A.  Because you are interested in the rate of complaints for a doctor, you should not impute based on complaints, as you would be essentially assuming an average rate of complaints for that doctor.

```{r}
library(VIM)

esdknn <- kNN(esdfinal[,-c(5)], k=3)
esdfinal[c(2,10,14),]
esdknn[c(2,10,14),]
```

B.  kNN should be based on taking all doctors of the same residency status and gender, then finding the two nearest in revenue, then taking the average number of visits for those two.

C.  Mean value imputation is based on taking all doctors of the same residency and gender, then taking the average number of visits.

```{r}
visits = by(esdfinal[,"visits"], INDICES = esdfinal[,c("residency", "gender")], FUN = mean, na.rm=TRUE)
visits
```

D.  Should say kNN.  Linear regression is also acceptable.  They should argue that the number of visits is related to the amount of work done (which can be measured by either revenue or hours).

#  3.  HTML pages. (25 pts)

In an HTML file such as those displayed on web browsers, the files are formatted through the form of HTML tags. For example, here is a very simple web html page:

```
<!DOCTYPE html>
<html>
<body>

<h1>My First Heading</h1>
<h2>This is heading level 2</h2>
<h3>This is heading level 3</h3>
<h3>This is another heading level 3</h3>
<h4>This is heading level 4</h4>
<h2>This is another heading level 2</h2>

<p>My first paragraph.</p>

</body>
</html>
```

Examples of tags are `<html>`, `<body>`, and `<h1>`, and they are always closed by something like `</html>`, `</body>`, and `</h1>`.
While there are standard tags, a page can define any tag desired. If the formatting of the tag is specified, then the web browser will use the tag, otherwise it is ignored.

You have a text file and you want to identify all headings in the document so that you can create a table of contents. You do not know how many levels of headings can be in the file, but you know it does not go more than 9 levels deep.

A.  How do you identify the beginning of the heading?
B.  How do you identify the heading level?
C.  How do you identify the end of the heading?
D.  Given that you know the character position (number) of every character in the file, and that you can identify the position of the results to (A) and (B), how do you specify the heading title?

ANSWER
------

A. Search for the string `<H#>` where `#` is a number from 1 to 9.

B. The value of `#` from the above.

C. Search for the next occurrence of the string `</H#>` where `#` is the string found in (A).  Note this answer needs to specify that it is looking for the same number `#` as in (A)

D. Starting character value is A + 4. Ending character is C - 1. You should eliminate starting and closing spaces, but that is not important.

#  4.  Kansas election results (25 pts)

```{r, echo=FALSE}
kansas <- read.delim(file = "2010-Kansas-General-Election-11-03.txt", header = TRUE, sep = "|")
kansashouse <- kansas[(substr(kansas$Contest, 1, 19)=="United States House"), c(1,2,3,4)]
racename <- function(contest){
  name <- paste("U.S.", substring(contest,15, 26), substring(contest, 40))
}
kansashouse$Contest <-  sapply(kansashouse$Contest, racename)
head(kansashouse, 20)
```                                         

The Kansas Secretary of States publishes election results in the format of Table 2, (which provides the first 20 entries). It breaks down voting for each county, contest (race) and candidate voted for. There are 105 counties in Kansas, and 4 house seats (which are not divided along county lines) with 13 candidates. However, what you are interested in is the total for each candidate in each contest across the state.

Note: the Contest is in the form *U.S. House of Rep 00#* where *#* represents the number of the district, e.g. *U.S. House of Rep 002* would indicate the 2nd Congressional District.

A.  Is this data in *wide* or *long* format?
B.  Provide the procedure to convert this data into a desired format then determine the winner of each U.S. House of Representatives Congressional District.  After each step, you should describe what the data looks like at each step.


ANSWER
------

A.  The data is in *long* format.  Each row shows a single vote tally.
B.  Reshape then aggregate
  i.  Use reshape to *cast* the data into a row for each county and race. So you will have a row for each county, and there will be columns for each candidate. Note that most candidate vote values will be 0 or NA.
  ii. *Aggregate* (taking the sum) votes by race and candidate (Note: each candidate is in only one race, so it is not necessary to say *race*).  You will have a row for each race, with only candidates for each race having non-zero values.
  iii.  *Select* the candidate with the maximum number of votes for each race.
