---
title: "IE 0015 Information Systems Spring 2020 Final"
output:
  html_document: default
  pdf_document: default
---


```{r setup, echo=FALSE, message = FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(phonenumber)
# replace my name with your name here and run all chunks before starting
yourname = "Jane Doe"
numbername = letterToNumber(yourname)
examseed = (as.numeric(gsub("-", "", numbername, fixed = TRUE)) %% 256)
print(paste("Name: ", yourname))
print(Sys.Date())
```

Instructions
=============

1. Enter your name in the code chunk above as the variable `yourname`
2. Run all chunks, then knit this file to see your exam with all of its specifications. (this exam is customized for you)
3. Read the agreement below and agree before starting.  Type your name in below the agreement to indicate your acceptance.
4. Open MS Teams into the IE 15 team. 
5. Take a picture of your working environment to include a clock (use time.gov if you need one).
6. (coding version) You have from 8:00 AM EDT to 1:00 PM (Noon) EDT to complete the exam.  I will be online in MS Teams from 8 to 12.
6. (Written version)  You have from 10 AM to 12 PM (NOON) EDT to complete the exam.
7. If you have a question, post that you have a question in the Team chat, and I will invite you to a private chat to answer your question. 
8. This is exam is open book and notes. You have access to 
    - your notes, , cheetsheets
    - the lecture notes, 
    - all homework  and labs, 
    -  the R for Data Science book website (https://r4ds.had.co.nz/), 
    - anything accessible through R Studio help section 
        -  Cheetsheets https://rstudio.com/resources/cheatsheets/
        -  Manuals - An Introduction to R
        -  All package documentation (vignettes and help files)

- Student agreement: 	"I have not received any unauthorized aid during this <assignment/quiz/exam> in accordance with the policies outlined above _________________."   Typing your name here indicates your agreement

## Bootstrap simulation COVID tracking data



```{r readcoviddata, echo=FALSE, message = FALSE, warning=FALSE}
datadirectory <- "."
#datadirectory <- "~/projects/infectiousdisease/covid-tracking-data"
#datadirectory <- "~/desktop/IE0015/COVID-19/covid-tracking-data"
statetracking <- read_csv(file.path(datadirectory, "states_daily_4pm_et.csv"),
              col_types = cols(date = col_date(format = "%Y%m%d"),
                  dateChecked = col_date(), 
                  death = col_number(),
                  negative = col_integer(), 
                  pending = col_number(),
                  positive = col_integer(), 
                  total = col_number(),
                  hospitalizedCumulative = col_integer(),
                  inIcuCurrently = col_integer(),
                  inIcuCumulative = col_integer()))
statetracking <- statetracking %>% 
  mutate(state = as.factor(state)) %>%
  replace_na(list(positive=0,
                  negative=0,
                  pending=0))

states <- sort(unique(statetracking$state))
summary(statetracking)
```