---
title: "IE 0015 Spring 2015 Midterm"
author: "Louis Luangkesorn"
date: "02/26/2015"
output: html_document
---


1.  CoNVO statement

Read the article on the changing of healthcare enrollments under the Accountable Care Act ("Obamacare") in its second year of enrollment. Take on the view of the Pennsylvania Department of Health. It is planning an overhaul of its public assistance of health care.
Develop a CoNVO on a data analysis similar to this article. Note: You should include in the Vision something that is not in the article.


http://www.nytimes.com/2015/02/27/upshot/high-rate-of-shopping-and-switching-in-obamacare-plans-is-a-good-sign.html?


Read the article on the changing distribution of jobs. Take on the view of the Allegheny County planning measures to encourage economic development.
Develop a CoNVO on a data analysis similar to this article. Note: You should include in the Vision something that is not in the article.

http://www.nytimes.com/2015/02/24/upshot/more-new-jobs-are-in-city-centers-while-employment-growth-shrinks-in-the-suburbs.html


2.  Data Analysis


```{r}
library(RSQLite)
con <- dbConnect(SQLite(), "../data/acs1yrPittsburgh01701.sqlite")
sql <- "SELECT *      
        FROM ss12hpa"
acsh <- dbGetQuery(con, sql)
sql <- "SELECT *      
        FROM ss12ppa"
acsp <- dbGetQuery(con, sql)
```

#  Get insurance status 
#  Get income status
#  Plots of income vs insurance type
#  Plots of 

```{r}


```



3.  USDA National Nutrient Database

https://www.ars.usda.gov/Services/docs.htm?docid=8964


Trainers at a fitness centers are interested in knowing the nutritional profiles of their clients. As part of their services, they ask new clients to keep a food diary with what they eat and the serving size.  So, the clients keep the following data for everything they eat during the week:

1.  Client id
2.  Dish
3.  Serving size

Using the USDA Nutrient Database (NDb), the trainers can then determine how many nutrients their clients have over the week.

The fitness center is interested in looking at the distribution of weekly calorie intake of their new members to compare with longer term members. Write out the sequence of steps needed to use the USDA Nutrient Database and the food diary to determine how many calories each of their clients eat during the week.


Dietitions working with a local grocery store are developing weekly menus for clients. The grocery is interested in the distribution of the weekly calorie content of the meals that their dietitians come up with.

1.  Client id
2.  Dish
3.  Serving size

Using the USDA Nutrient Database (NDb), the trainers can then determine how many nutrients their clients have over the week.

Write out the sequence of steps needed to use the USDA Nutrient Database and the food diary to determine the distribution of number of calories being recommended by the grocery dietitions to their clients.


Dietitions working at a nursing home are developing weekly menus for residents. The Nursing home are interested in the number and type of calories residents are taking in every week.  The weekly menus are stored in a table that for each dish and each meal contains the following:

1.  Client id
2.  Dish
3.  Serving size

Using the USDA Nutrient Database (NDb), the dietitions can then determine how many calories and if they come from protein, fat, or cholestorol.

Write out the sequence of steps needed to use the USDA Nutrient Database and the weekly menus to determine the distribution of number of calories by type for their residents.