---
title: 'Lab 9: Regression'
author: "IE 0015 Information Systems Engineering"
date: "March 28/29, 2019"
output:
  pdf_document: default
  html_document: default
---

This lab will use the Titanic data. You should modify the `read.csv` function call to correspond to your system.

```{r loadlibraries,  warning=FALSE}
options(digits = 3, scipen=4)
library(ggplot2, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(magrittr)
library(modelr, quietly = TRUE, warn.conflicts = FALSE)

titanic <- read.csv("../data/titanic3.csv",
                    header=TRUE, sep=",")
# From Lab on working with characters and regular expressions
getTitle <- function(data) {
  title.dot.start <- regexpr("\\,[A-Z ]*\\.", data$name, TRUE)
  title.comma.end <- title.dot.start+ attr(title.dot.start, "match.length")
  data$Title <- substr(data$name, title.dot.start+2, title.comma.end-1)
  return (data$Title)
}  
titanic$Title <- getTitle(titanic)
unique(titanic$Title)  
```


# 1. Multiple linear regression (4 pts)

You think that the age is related to the following:

  i.  Sex
  ii.  Pclass

a.  Remove lines in the data that do not have an age
b.  Develop a multiple linear regression model for the age based on these characteristics.
c.  Display the summary and the analysis of variance tables for this model. 
d.  Display diagnostics plots. Discuss the assumptions of regression models.
e.  Do you think this is an adequate model?
  
**ANSWER**


a-b-c. Model
```{r}
#a 
titanic <- titanic %>%
  filter(!is.na(titanic$age))
#b
age.lm <- lm(age ~ sex + pclass, data=titanic)
#c
summary(age.lm)
anova(age.lm)
```
d. Diagnostic plots
```{r}
par(mfrow=c(2,2)) 
plot(age.lm)
par(mfrow=c(1,1))
```

-  Normality (looks pretty good, only breaks off the 45 degree line near the top of the distribution)
-  Error distribution seems independent looking at the Residuals vs fitted and scale-location graphs (i.e. the standardized residuals cover the same range)
-  Constant variance (only changes as for passengers that are  very young or very old)
-  Bonus: Cook's distance does not show the outer boundaries in the plot (i.e. the boundaries are out of the plot limits), so it passes.

e. It is adequate (i.e. all tests pass)  You could comment that there are only 6 predicted values, because of the combination of gender and pclass, so it seems too discrete while ages are continuous.

# 2. Adding to the model. (4 pts)

There are two variables that are generally present but are not currently in the model: fare and Title. 

a. Plot the residuals of your model against each of these and choose which value should be added to the model.
b. Create the linear regression model that adds the new variable and print the summary, anova table, and diagnostic plots.
c. Compare the new model to the original.

**ANSWER**

a. Compare the residuals against the Title, then fare.

```{r}
titanic$lmresiduals <- age.lm$residuals
ggplot(titanic) + geom_boxplot(aes(Title, lmresiduals)) + 
  ggtitle("Age prediction errors by social title") +ylab("Residuals on age prediction")
```

```{r}
ggplot(titanic) + geom_point(aes(fare, lmresiduals)) + 
  ggtitle("Age prediction errors by fare") +ylab("Residuals on age prediction")
```

The residuals for each fare seem to have the same range for most values of the fare.  But their are large differences in the boxplots when plotting against the title, indicating that the title is informative, so add the title to the regression model.

b.  

```{r}
age.lm2 <- lm(age ~ sex + pclass + Title, titanic)
summary(age.lm2)
anova(age.lm2)
```

Diagnostic plots

```{r}
par(mfrow=c(2,2))
plot(age.lm2)
par(mfrow=c(1,1))
```


c. 
There are more varied predicted values of age. Many of the residuals (young and old) are very small because of few samples. For those with many samples, the residuals are smaller than before, and about even.

- Note that Cook's distance is still within the dotted lines. (the lower bound may not show up)
- QQ plot looks better at the upper end of the range.
- Sum of squares residuals is much smaller. 127776 vs 177602
- Adjusted R^2 increases from 0.18 to 0.401
- Plot of errors by title shows reduced residuals (note that y-axis is much smaller)

```{r}
titanic$lmresiduals2 <- age.lm2$residuals
ggplot(titanic) + geom_boxplot(aes(Title, lmresiduals2)) + 
  ggtitle("Age prediction errors by social title - 2") +ylab("Residuals on age prediction")
```


3.  Use the final model from problem 2 to predict the ages of the individuals listed below. (2 points)  

Hint: look up `predict` in help.

Note: remember to explain the output.

```{r predictpassenger}
predicttitanic <- titanic[c(13, 51), c("pclass", "name", "sex", "sibsp", "embarked", "Title")]
predicttitanic
```


**ANSWER**

```{r}
predict(age.lm2, predicttitanic)
```

The predicted age for Mme. Auburt is 51, the predicted age for Master Carter is 15.