---
title: 'Lab 3:  Aggregation'
author: "IE 0015 Information Systems Engineering"
date: "January 24/25, 2019"
output:
  html_document: default
  pdf_document: default
---


Load the data set from the National Survey of Family Growth from CourseWeb. (i.e. download the data and modify the below read_csv function call appropriately)
```{r loadlibraries, warning=FALSE}
library(dplyr)
library(readr)
library(magrittr)
# NSFG data in Courseweb
FemPreg <- read_csv("../data/nsfgFemPreg.csv",
                    col_names=TRUE)
FemPreg$BIRTHORD <- as.factor(FemPreg$BIRTHORD)
FemPreg$PRGLNGTH <- as.integer(FemPreg$PRGLNGTH)
```

```{r}
# level of education
education <- data.frame(
          HIEDUC=c(5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15),
          education = c("9TH GRADE","10TH GRADE",
             "11TH GRADE", "12TH GRADE",
             "HIGH SCHOOL", "SOME COLLEGE",
             "ASSOCIATE", "BACHELOR",
             "MASTER", "DOCTORATE", "PROFESSIONAL DEGREE"))
# informal marital status
marital <- data.frame(RMARITAL=c(1, 2, 3, 4, 5, 6),
                      marital = c("MARRIED",
                                 "LIVING WITH PARTNER",
                                 "WIDOWED",
                                 "DIVORCED",
                                 "SEPARATED",
                                 "NEVER BEEN MARRIED"))
```

1. [2] The HIEDUC and RMARITAL fields currently have codes.  The `education` and `marital` data frames contain the definitions behind those codes based on the NSFG code book.  Use `left_join` to add a columns to `FemPreg` to include the actual values of level of education and marital status.

**ANSWER**

```{r}
FemPreg <- FemPreg %>% left_join(education, by = "HIEDUC")

FemPreg <- FemPreg %>% left_join(marital, by="RMARITAL")
```

2. [2]  Using the female pregnancy data, create a table  that shows the distribution of respondents by level of education. Note that there is more than one way to approach this.  Use the dplyr functions.


## ANSWER
```{r}
FemPreg %>% group_by(education) %>%
  summarize(count=n())
```

# alternative solutions:
```{r}
as.data.frame(table(FemPreg$education))
edu<-FemPreg$education
edudf<-as.data.frame(summary(edu))

as.data.frame(summary(subset(FemPreg,select=c("education"))))
```

3. [3]  Find the mean of number of weeks of breastfeeding  by level of education of the mother. (Note that the code above merges the ACS coded level of education with its meaning.) 
  i.  Create a data frame that only includes women who have completed breastfeeding (hint: remove those for which are not applicable, still breastfeed, or never breastfeed based on the BFEEDWKS field, a '!' indicates  The Female pregnancy data codebook identifies the codes used in the `BFEEDWKS` variable.)
  ii.  Take the mean number of weeks of breastfeeding by education (of mother) 

**ANSWER**
```{r}
FemPreg %>% filter(!BFEEDWKS %in% c(NA, 994, 995)) %>%
  group_by(education) %>%
    summarise(avgweeks = mean(BFEEDWKS))
```

# Alternative solutions:

```{r}
completebf <- FemPreg[!(FemPreg$BFEEDWKS %in% c(NA, 994, 995)),]
completebf$BFEEDWKS <- as.numeric(completebf$BFEEDWKS)
aggregate(completebf[,c('BFEEDWKS')],by=list(completebf$education),mean,na.rm=TRUE)
```
```{r}
# tapply mostly works on vector, so it will need two lines, each for one vector. 
tapply(completebf$BFEEDWKS,completebf$education,mean)
```

4. [3]  Find the average number of weeks of breastfeeding and weeks of maternity leave based on education and informal marital status (`education` and `marital`). Note: remember that to ignore 'NA' in the mean, use the option `na.rm=TRUE`  e.g. `mean(vector, na.rm=TRUE)`

**ANSWER**
```{r}
completebf %>% group_by(education, marital) %>%
  summarise(avgweeks=mean(BFEEDWKS, na.rm=TRUE))
```

**Alternative**

```{r, warnings=FALSE}
aggregate(completebf[,c('BFEEDWKS')],by=list(completebf$education, completebf$marital),mean,na.rm=TRUE)
```


Remark:   
1) When there is Error message like "Error: could not find function "cast" ", possible reason is that you haven't 
loaded the library or haven't installed it. Installing or loading the library should fix the problem.       
