---
title: "Lab 6: Complex graphs"
author: "IE 0015 Information Systems"
date: "February 2020"
output:
  html_document: default
  pdf_document: default
---

```{r loadlibraries, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(magrittr)
library(RColorBrewer)
```
```{r, loaddata}
FemPreg <- read.csv("../data/nsfgFemPreg.csv",
                    header=TRUE, sep=",")
```


1.  [5] One question when looking at population demographics is if all parts of the population have the same distribution. 

    a.  Using the Female Pregnancy Data from the National Survey of Population Growth, create graphs that shows the distribution of the birthweight of newborn babies, separated by race (Note: use HISPRACE for this purpose).  Use at least three different type of plots (geom_ or stat_ commands from ggplot, different ways of displaying multiple categories on a single plot).
    b.  Explain which of the three you used is best.
    c.  Does race play a factor in the distribution of birthweight?
    
Note: 

  1.  You will need both a way to display distributions, and a way to show each race in a single plot. 
  2.  Race should be identified in the plot as a label, not as a number. (which is how it is presented in the Female Pregnancy data table.)
  3. The code below converts numeric values into their codes (see Lecture 3: Data in R). You will need to do something similar later.
  4. Because you are getting distributions by race, you should convert counts into percentages using `y=..density..` within the `aes()` function. 

```{r}
raceorigin <- as.data.frame(
           rbind(c("1", "Hispanic"),
                 c("2", "Non-Hispanic White"),
                 c("3", "Non-HIspanic Black"),
                 c("4", "Non-HIspanic Other"))
           )
colnames(raceorigin) <- c("HISPRACE", "raceorigin")
FemPreg$HISPRACE <- as.factor(FemPreg$HISPRACE)
FemPreg <- FemPreg %>% 
  left_join(raceorigin)
FemPreg$BIRTHWG_LB1 <- as.numeric(FemPreg$BIRTHWGT_LB1)
FemPreg <- FemPreg %>% filter(BIRTHWGT_LB1<98)
```
**ANSWER**

Look for each race to be separated.

Note: only partial credit for facet_wrap/facet_grid since they are not single graphs.
```{r}
FemPreg <- FemPreg %>% mutate(birthweight = BIRTHWGT_LB1 + BIRTHWGT_OZ1/16.0)
ggplot(FemPreg) + geom_histogram(aes(birthweight, y=..density..), binwidth=1) +facet_grid(raceorigin ~.) +
  xlab("Birthweight (lbs)")
```
Should be either stacked, dodge, 

```{r}
ggplot(FemPreg) + geom_histogram(aes(birthweight, , y=..density.., fill=raceorigin), binwidth=1)+
  xlab("Birthweight (lbs)")
```


```{r}
ggplot(FemPreg) + geom_histogram(aes(birthweight, , y=..density.., fill=raceorigin), binwidth=1, position="dodge")+
  xlab("Birthweight (lbs)")
```



```{r}
ggplot(FemPreg) + geom_density(aes(birthweight, y=..density.., fill=raceorigin), binwidth=1, position="stack")+
  xlab("Birthweight (lbs)")
```

2.  [5] The American Academy of Pediatrics recommends that infants are breastfed exclusively for about the first 6 months and that breastfeeding continue for at least 12 months.  Plot the average duration of breastfeeding (including none) according to both race (HISPRACE) and level of education of the mother (HIEDUC.  Make sure that the plot can be understood when printed in black and white and that the axis names are meaningful and legible. (i.e. translate the numerical values to the descriptions)

    a.  What kind of plot will you choose and why?
    b.  Create the plot using ggplot.  Make sure you have meaningful and legible axis titles.
    c.  Discuss if race or education shows a difference in the distributions in breastfeeding period.


Notes: 

  1. Use BFWEEKS indicate length of breastfeeding.  
  2. Think about what NA means for this variable.

  
```{r}
FemPreg$HIEDUC <- as.ordered(FemPreg$HIEDUC)
highesteducation <- as.data.frame(
                     rbind(c("5", "9TH GRADE OR LESS"),
                            c("6", "10TH GRADE"),
                            c("7", "11TH GRADE"),
                            c("8", "12TH GRADE, NO DIPLOMA"),
                            c("9", "HIGH SCHOOL GRADUATE OR GED"),
                            c("10", "SOME COLLEGE BUT NO DEGREE"),
                            c("11", "ASSOCIATE DEGREE"),
                            c("12", "BACHELOR'S DEGREE"),
                            c("13", "MASTER'S DEGREE"),
                            c("14", "DOCTORATE DEGREE"),
                            c("15", "PROFESSIONAL DEGREE")
                     )
           )
colnames(highesteducation) <- c("HIEDUC", "highesteducation")
FemPreg <- FemPreg %>% 
  left_join(highesteducation)
```

**ANSWER**

Note: has to have every combination of Race and Education represented somewhere.


```{r}
breastfeeding <- FemPreg %>% filter(BFEEDWKS < 100)
ggplot(breastfeeding) + geom_bar(aes(x=highesteducation)) +facet_grid(.~raceorigin) + 
  theme(axis.text.x = element_text(angle=90)) +  xlab("Highest education level achieved") 
```

Note: Boxplots do include medium instead of mean, so this works.  

```{r}
breastfeeding <- FemPreg %>% filter(BFEEDWKS < 110)
ggplot(breastfeeding) + geom_boxplot(aes(x=highesteducation, y=BFEEDWKS)) +facet_grid(.~raceorigin) + 
  theme(axis.text.x = element_text(angle=90)) +  xlab("Highest education level achieved") 
```