---
title: 'Lab 7: TVM and Advanced programming'
author: "Louis Luangkesorn"
date: "February 21/22, 2020"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=2)
```

Note: Knit this document to HTML before starting. There are equations that will render.

# Investment projects

As a manager with an investment budget, you need to allocate your funds across projects to get the highest overall rate of return. Each investment has an associated discount rate (based on the loan that could be obtained for it), up front investment required, and results in an annual cash flow for a specified period of years.  This assignment will build a function that will help choose which combination of projects to pick.

```{r}
project <- data.frame(project = c('A', 'B', 'C', 'D', 'E'),
                      r = c(0.10, 0.12, 0.08, 0.08, 0.12),
                      investment = c(3, 4, 5, 3, 3),
                      paymentvalue = c(1, 1, 2, 1.5, 1),
                      years = c(5, 8, 5, 4, 6)
                      )
project
```
# Time value of money

The time value of money relates the present value (PV) to future value (FV) based on an interest rate (r) and the time horizon (t).

$$PV = \frac{FV}{(1+r)^t}$$

$$FV = PV {(1+r)^t}$$

So the expression $1/(1+r)^t$ is the discount factor, of the present value of \$1 received in year $t$.

# 1. Write time value of money functions [2 pts]

a. Draw a flow chart that calculates present value (PV) given $FV, r$, and $t$.
b. Write a function in R to calculate PV given $FV, r,$ and $t$. Note for this and (b), you need to have comments in the function that specify what the parameters are.
c. Write a function in R to calculate FV given $PV, r,$ and $t$.
d. Use the PV function to solve for the present value of \$1,000 to be paid in twenty-five years when the annual interest rate is 8.53\%.
e. Use the FV function to calculate the future value of \$1,200 that is put in the bank now in two years when the annual interest rate is 8\%.

**ANSWER**

```{r tvmfunctions}
# (a)
PVM <- function(FV, r, t){
  # FV - Future value
  # r - interest rate
  # t - time periods
  PVM <- FV / ((1+r)^t)
  PVM
}
# (b)
FVM <- function(PV, r, t){
  # PV - Present value
  # r - interest rate
  # t - time periods
  FVM <- PV * ((1+r)^t)
  FVM
}

# (c)
PV1000 <- PVM(1000, 0.0853, 25)
# (d)
FV1200 <- FVM(1200, 0.08, 2)
print(c(PV1000, FV1200))
```

# 2.  Bond valuation [4]

a.  Write the flow chart that finds the present value of a bond. A bond consists of a series of future payments, a final payout at maturity,  and the current interest rate. Note: you may assume that the PVM function use the PVM function you have created in (1) is available (i.e. it takes a single box).  
b.  Create a function that finds the present value of a bond (i.e. implements (2a).  In the comments of the function you need to define what the parameters are. (i.e. comments so you know how to use the function in the future.)
c.  Using your bond function, find the present values of the following bond:
    (i)  Payments of \$100 at the end of every year for three years, payout at maturity at the end of year 3 of \$1000 (note that one of the annual payments occurs simultaneously), and a rate of return of 10\%.

**ANSWER**

```{r bondpayments}
bondvalue <- function(payment, npayments, payout, maturitytime, r){
  # payment are the regular payments,
  # npayments are the number of regular payments over the life of the investment
  # payout is the value at maturity
  # maturitytime is the number of periods until the end of life of the investment 
  # r is the interest rate.
  paymentvalue <- sum(PVM(payment, r, 1:npayments))
  payoutvalue <- PVM(payout, r, maturitytime)
  PV <- paymentvalue + payoutvalue
  PV
}
payment = 100
npayments = 3
maturitytime = 3
r = 0.1
payout = 1000
print(bondvalue(payment, npayments, payout, maturitytime, r))
```

# Value of projects [4]

a.  Calculate the present value of each of the projects from the table above.  Use your bond function and present value functions as appropriate. 
b.  What is the best combination of investments for an \$8 million budget?

Note: investments (your payments) are modeled as a negative payout.
Note: Projects do not usually have a payout at the end (you invest, and you get an ongoing return, but you would only get an end of project payout if you could sell the project at salvage value.)
Note: *Best* is defined as highest total present value of the project.

**ANSWER**
```{r projectvalue}
valueA <- bondvalue(payment = project$paymentvalue[1], npayments = project$years[1], payout = 0, maturitytime = project$years[1], r = project$r[1]) - project$investment[1]

valueB <- bondvalue(payment = project$paymentvalue[2], npayments = project$years[2], payout = 0, maturitytime = project$years[2], r = project$r[2]) - project$investment[2]

valueC <- bondvalue(payment = project$paymentvalue[3], npayments = project$years[1], payout = 0, maturitytime = project$years[3], r = project$r[3]) - project$investment[3]

valueD <- bondvalue(payment = project$paymentvalue[4], npayments = project$years[4], payout = 0, maturitytime = project$years[4], r = project$r[4]) - project$investment[4]

valueE <- bondvalue(payment = project$paymentvalue[5], npayments = project$years[5], payout = 0, maturitytime = project$years[5], r = project$r[5]) - project$investment[5]
c(valueA, valueB, valueC, valueD, valueE)
```

So, the best investment will be the largest total for investments adding up to 8. So projects C and D.

