---
title: "Lab 7: Complex visualizations"
author: "IE 0015 Information Systems"
date: "February 23, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r loadlibraries, warning=FALSE}
library(ggplot2)
library(dplyr)
library(magrittr)
```
```{r, loaddata}
FemPreg <- read.csv("../data/nsfgFemPreg.csv",
                    header=TRUE, sep=",")
```


1.  [3] One question when looking at population demographics is if all parts of the population have the same distribution. Using the Female Pregnancy Data from the National Survey of Population Growth, create a graph that shows the distribution of the birthweight of newborn babies, separated by race (Note: use HISPRACE for this purpose).  

Note: 

  1.  You will need both a way to display distributions, and a way to show each race in a single plot. 
  2.  Race should be identified in the plot as a label, not as a number. (which is how it is presented in the Female Pregnancy data table.)
  3. The code below converts numeric values into their codes (see Lecture 4: Data in R). You will need to something similar later.
  4. Because you are getting distributions by race, you should convert counts into percentages using `y=..density..` within the `aes()` function. 

```{r}
raceorigin <- as.data.frame(
           rbind(c("1", "Hispanic"),
                 c("2", "Non-Hispanic White"),
                 c("3", "Non-HIspanic Black"),
                 c("4", "Non-HIspanic Other"))
           )
colnames(raceorigin) <- c("HISPRACE", "raceorigin")
FemPreg$HISPRACE <- as.factor(FemPreg$HISPRACE)
FemPreg <- FemPreg %>% 
  left_join(raceorigin)
```

2.  (6) The American Academy of Pediatrics recommends that infants are breastfed exclusively for about the first 6 months and that breastfeeding continue for at least 12 months.  Plot the average duration of breastfeeding (including none) according to both race (HISPRACE) and level of education of the mother (HIEDUC.  Make sure that the plot can be understood when printed in black and white and that the axis names are meaningful and legible. (i.e. translate the numerical values to the descriptions)

Notes: 

  1. Use AGEQTNURN1 indicate length of breastfeeding.  
  2. Think about what NA means for this variable.
  3. Aggregations and summaries were covered in Lecture 6: Aggregation

  a.  What kind of plot will you choose and why?
  b.  Create the plot using ggplot.  Make sure you have meaningful and legible axis titles.
