---
title: 'Lab 7: Character Strings'
author: "IE 0015 Information Systems"
date: "02/16/2016"
output: pdf_document
---

```{r loadtitanic, echo=FALSE}
titanic <- read.csv("../data/titanictrain.csv")
```
1.  (3 pts) The Titanic accident was in a time when people were identified by their titles. In the Titanic data, How do you identify the title (Mr., Mrs., Miss. etc) of a person.

ANSWER

Look in the name field, find the comma after the last name.
The title starts two spaces after the comma.
The title is completed with a period.

2.  (5 pts) Read in the titanic data and extract the title of each person.  Make a list of unique titles.
  (hint, how do you find the first letter? how do you know the title is complete? Note: The Countess is a special case, don't worry about getting her territory right, just her title.  Note 2:  in a regular expression, you need to escape the punctuation with a backslash, and you need to escape the backslash with another backslash (i.e. two backslashes before a punctuation mark.)
  
  
ANSWER

```{r}
getTitle <- function(data) {
  title.dot.start <- regexpr("\\,[A-Z ]*\\.", data$Name, TRUE)
  title.comma.end <- title.dot.start+ attr(title.dot.start, "match.length")-1
  data$Title <- substr(data$Name, title.dot.start+2, title.comma.end-1)
  return (data$Title)
}  
titanic$Title <- getTitle(titanic)
unique(titanic$Title)  
```

3.  (2 pts) How many passengers have each title? (hint: declare the list of titles to be a factor and take the summary)

ANSWER

```{r}
titanic$Title <- factor(titanic$Title)
summary(titanic$Title)
```
