complex graphs with ggplot2 Answer
===========================
author: "IE 0015 Information Systems"
date: February 2017

```{r loadlibraries, warning=FALSE}
library(ggplot2)
library(dplyr)
library(mgcv)
```
```{r, loaddata}
FemPreg <- read.csv("../data/nsfgFemPreg.csv",
                    header=TRUE, sep=",")
```


1.  [3] One question when looking at population demographics is if all parts of the population have the same distribution. Using the Female Pregnancy Data from the National Survey of Population Growth, create a graph that shows the distribution of the birthweight of newborn babies, separated by race (Note: use HISPRACE for this purpose).  Note: you will need both a way to display distributions, and a way to show each race in a single plot. Race should be identified in the plot as a label, not as a number. (which is how it is presented in the Female Pregnancy data table.)

Answer:
```{r}
ggplot(FemPreg[FemPreg$BIRTHWGT_LB1<98,], aes(as.numeric(BIRTHWGT_LB1), ..density..,  linetype=as.factor(HISPRACE), fill=as.factor(HISPRACE))) + 
  geom_histogram(center=2, bins=10, position="dodge") +
  scale_x_discrete(limits=c(2,12), breaks=c(2,4,6,8,10,12), name="Birthweight (lbs)") +
  scale_y_continuous(labels= scales::percent)

```

2.  (6) The American Academcy of Pediatrics recommends that infants are breastfed exclusively for about the first 6 months and that breastfeeding continue for at least 12 months.  Plot the duration of breastfeeding (including none) according to both race (HISPRACE) and level of education of the mother.
  a.  What kind of plot will you choose and why?
  b.  Create the plot using ggplot.  
