---
title: 'Lab 9:  Aggregation'
author: "IE 0015 Information Systems Engineering"
date: "03/17/2015"
output: html_document
---

```{r loadlibraries, warning=FALSE}
library(reshape)
library(plyr)
FemPreg <- read.csv("../data/2006_2010_FemPreg.csv",
                    header=TRUE, sep=",")
FemPreg$BIRTHORD <- as.factor(FemPreg$BIRTHORD)
FemPreg$PRGLNGTH <- as.integer(FemPreg$PRGLNGTH)
```

```{r}
# level of education
FemPreg$education <- factor(mapvalues(FemPreg$HIEDUC,
          from=c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"),
          to= c("9TH GRADE","10TH GRADE",
             "11TH GRADE", "12TH GRADE",
             "HIGH SCHOOL", "SOME COLLEGE",
             "ASSOCIATE", "BACHELOR",
             "MASTER", "DOCTORATE", "PROFESSIONAL DEGREE")), ordered=TRUE)
# informal marital status
FemPreg$marital <- factor(mapvalues(FemPreg$RMARITAL,
                                    from=c(1, 2, 3, 4, 5, 6),
                                    to = c("MARRIED",
                                           "LIVING WITH PARTNER",
                                           "WIDOWED",
                                           "DIVORCED",
                                           "SEPARATED",
                                           "NEVER BEEN MARRIED")), ordered=TRUE)
```

1.  Using the female pregnancy data, create a data frame that shows the distribution of respondents by level of education.


ANSWER
```{r}
as.data.frame(table(FemPreg$education))
# Optional solution:
edu<-FemPreg$education
edudf<-as.data.frame(summary(edu))

as.data.frame(summary(subset(FemPreg,select=c("education"))))
```

2.  Find the mean of number of weeks of breastfeeding and the weeks of maternity leave taken by level of education. 
  i.  Create a data frame that only includes women who have completed breastfeeding (hint: remove those for which are not applicable, still breastfeed, or never breastfeed based on the BFEEDWKS field, use the !(FemPreg$BFEEDWKS %in% c( . . .)) to select from the data frame.)
  ii.  take the mean by factor (education) (note: to ignore NA's, use the argument `na.rm=TRUE` for the colMeans)

ANSWER

```{r}
completebf <- FemPreg[!(FemPreg$BFEEDWKS %in% c(NA, 994, 995)),]
completebf$BFEEDWKS <- as.numeric(completebf$BFEEDWKS)
completebf$MATLEAVE <- as.numeric(completebf$MATLEAVE)
by(completebf[,c('BFEEDWKS', 'MATLEAVE')], completebf$education, colMeans, na.rm=TRUE)
# Optional solutions:
aggregate(completebf[,c('BFEEDWKS', 'MATLEAVE')],by=list(completebf$education),mean,na.rm=TRUE)
# tapply mostly works on vector, so it will need two lines, each for one vector. 
tapply(completebf$BFEEDWKS,completebf$education,mean)
tapply(completebf$MATLEAVE,completebf$education,mean)

```


3.  Using the `melt` then `cast` functions from the `reshape` library, find the average number of weeks of breastfeeding and weeks of maternity leave based on education and informal marital status (`education` and `marital`)

ANSWER
```{r}
library(reshape)
bfm <- reshape::melt(completebf, c('education', 'marital'), c('BFEEDWKS', 'MATLEAVE'))
reshape::cast(bfm, education+marital~variable, mean, na.rm=TRUE)

# Optional solution:
library(reshape2)
bfm <- reshape2::melt(completebf, c('education', 'marital'), c('BFEEDWKS', 'MATLEAVE'))
reshape2::dcast(bfm, education+marital~variable, mean, na.rm=TRUE)
```
Remark:   
1) When there is Error message like "Error: could not find function "cast" ", possible reason is that you haven't 
activate the library or haven't install it. Install or activate the library should fix the problem.       

2) Melt function is used for preparing data for cast function. For the melt command above, it will extract 4 columns from "completebf". For each row of the data frame, the melt command will create two new rows containing the information from the 4 blocks being extracted. For example, a row containing "BACHELOR,MARRIED,4,10" in "completebf"" will be represented by 2 rows "BACHELOR,MARRIED,BFEEDWKS,4" and "BACHELOR,MARRIED,MATLEAVE,10" in the melted dataframe.   


